---
title: "JTRCI: Jacobson-Truax & reliable change indices"
output:
  github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r include=FALSE}
knitr::opts_chunk$set(dpi=300, out.height ="450px", fig.asp = .71, fig.pos = "h")
```

```{r warning=FALSE, include=FALSE}
require(ggplot2)

source('~/OneDrive - psychology.su.se/AW r functions and stuff/JT-RCI/R/JT-RCI.R')
source('~/OneDrive - psychology.su.se/AW r functions and stuff/JT-RCI/R/plotJT.R')
source('~/OneDrive - psychology.su.se/AW r functions and stuff/JT-RCI/R/plotRCI.R')
```

```{r include=FALSE}
df <- read.csv("data/J&T 1991 table2 DAS data.csv")

JTRCI(data = df, 
      ppid = "ppid", 
      pre = "pre", 
      post = "post",  
      reliability = .84, 
      indextype = "JT", 
      JTcrit = "A", 
      higherIsBetter = T,
      table = F,
      plot = F)
```

Example plot using the Dyadic Adjustment Scale (DAS) data provided in table 2 by Jacobson & Truax (1991). 

```{r echo=FALSE}
plotJT(xlab = "DAS pre", ylab = "DAS post")
```

*notice that this is a 'higher is better' measure*

##
to install JTRCI, first install package devtools from CRAN - either through the 'install packages' interface (in for instance Rstudio) or using the following line:
```{r eval=FALSE}
install.packages("devtools")
```

then install JTRCI from github:
```{r eval=FALSE}
devtools::install_github("AWKruijt/JT-RCI")

library(JTRCI)
```

#
JRTCI() is the main function in this package: 

|    JTRCI (data = NA, pre = NA, post = NA, ppid = NA, group = NA,    
|                reliability = NA, higherIsBetter = F, indextype = "JT", JTcrit = "auto",     
|                normM = NA, normSD = NA, dysfM = NA, dysfSD = NA,     
|                plot = T, table = T, ...) 

It requires data in wide format (one row per individual) and an estimate of the measure reliability to be given through parameter 'reliability ='. 

The reliability estimate can be based on norm data. Alternatively, the internal reliability of the measure observed in the current data can be used. In that case, the user will first need to obtain their estimate of choice (alpha, omega, x-random splits splithalf reliability) using the raw (item level) data. 

Obtaining reliable change indices is pretty straightforward: run the function with parameter 'indextype = "RCI"'. The JTRCI function's default setting is to obtain Jacobson-Truax indices ('indextype = "JT"'). The function is designed to provide guidance through the various choices a researcher has to make when applying the Jacobson-Truax method. It is recommended to pay attention to the various output messages.  

#
When plot = T (default), the JTRCI() function will call on either the JTplot() or RCIplot().

The JTplot()/RCIplot() functions take the following parameters:

|    plotJT(data = JTRCIdf, useGroups = F, facetplot = F, addJitter = F, 
|              xlab = "pre", ylab = "post", plottitle = "Jacobson-Truax plot", 
|              addInfoLegend = c("yes", "classcounts", "JTcrit", "no"))

These parameters can also be passed directly into a call on JTRCI() (see examples below)

#

An online dashboard using essentially the same code is available at: https://awkruijt.shinyapps.io/JTRCI_dashboard/
It allows the user to upload their own data (as a .csv file) or to test things using generated/mock data. JTRCI plots and dataframes can subsequently be downloaded. 

An example of a more advanced plot (with multiple follow-ups) based on the JTRCI-outputs can be found here: https://awkruijt.netlify.com/plotposts/longitudinaljtrci

#
#### examples: 

```{r }
# generate some random data:
df <- cbind.data.frame("ppid" = seq(1:64), 
                       "pre" = rnorm(64, 65, 8), 
                       "post" = c(rnorm(32, 40, 8), rnorm(32, 45, 8)),
                       "group" = rep(c("treatment", "control"), each = 32))
```


Obtain and plot Jacobson-Truax indices using parameter JTcrit = "auto"
```{r}
# The function will determine which criterion to use based on the available information 
# f.i. when no (healthy or dysfunctional) norm values are provided, the function will  
# return criterion A using the baseline distribution as the 'dysfunctional distribution':

JTRCI(data = df, 
      ppid = "ppid", 
      pre = "pre", 
      post = "post",  
      group = "group", 
      reliability = .8, 
      indextype = "JT", 
      JTcrit = "auto")
```

Obtain reliable change indices by setting parameter indextype = "RCI":
```{r}
# obtain reliable change indices 
# also disable plotting with plot = F:
JTRCI(data = df, pre = "pre", post = "post", group = "group", ppid = "ppid",
      reliability = .8, indextype = "RCI",  table = F, plot = F)
```

Subsequently plot the RCI with a separate call to plot_RCI():
```{r}
# plot the reliable change indices obtained in the previous chunk - set x y and plot labels:

plotRCI(xlab = "score pre", ylab = "score post", plottitle = "my reliable change plot")
```

Parameters for the plot_JT()/plot_RCI() functions can also be passed directly in the JTRCI() function:
```{r}
# obtain and plot Jacobson-Truax indices using criterion B. 
# add jittering of the points with addJitter = T

JTRCI(data = df, pre = "pre", post = "post",  group = "group", ppid = "ppid",
      reliability = .8, indextype = "JT", JTcrit = "B",
      normM = 30, normSD = 5, 
      useGroups = T, addJitter = T, plottitle = "JT indices (crit B):")
```

```{r}
# obtain Jacobson-Truax indices using criterion C.
# plot with parameter facetplot = T - the plot function can handle up to 5 groups

JTRCI(data = df, pre = "pre", post = "post",  group = "group", ppid = "ppid",
      reliability = .8, indextype = "JT", JTcrit = "C",
       normM = 30, normSD = 5, dysfM = 60, dysfSD = 5, 
      facetplot = T, plottitle = "JT indices by group:")
```

It is important to note that when assessing/plotting Jacobson-Truax indices for multiple groups, the function combines the groups into one group for the calculations of the RCI (when using baseline characteristics as the 'dysfunctional distribution), or uses one single set of inputs dysfM and dysfSD as characteristic of the dysfunctional population. I.e. it assumes that the participants all originate from the same (sub)population and were randomly assigned to the different groups/conditions.

Generally speaking (regardless of whether there are multiple conditions), Jacobson-Truax indices are designed for use with participants selected so that they would be expected to score high (or low in case of a higher-is-better measure) on the measure assessed. For a control group in which most/all participants score already within the 'functional' (norm) range the traditional Jacobson-Truax classification labels make little sense: i.e. someone who scored in the healthy range at both the pre- and the post-assessment would be classified as 'recovered' without an actual recovering happening. The more 'simple' reliable change indices can be used for comparison of groups preselected to (likely) differ at baseline on the measure assessed. 

I've also taken the liberty of adding a new label to the Jacobson-Truax classification to indicate a sequence of events seen in the data that this code was originally developed for: participants who dropped below the JT-criterion at post, but did not actually show a pre-post change sufficiently large to be classified as reliable change (i.e. they scored already close to the JT-criterion value at basline). Such participants are labeled 'non-reliably recovered'. 